\ourTitle

### 2022-06-08 paper/manuscript.tex
JPC: Editing.
JPC: Editing almost done.

\begin{figure*}[ht]
\centering
\includegraphics[width=0.4\linewidth]{../plots/even_CANTOR.png}
\includegraphics[width=0.4\linewidth]{../plots/anbn_CANTOR.png}
\includegraphics[width=0.4\linewidth]{../plots/anbncn_CANTOR.png}
\includegraphics[width=0.4\linewidth]{../plots/x+f(x)_CANTOR.png}
\caption{Cantor plots for the even, $\mathtt{a}^n\mathtt{b}^n$,
	$\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and $\mathtt{x+f(x)}$ processes.
	Each point $(x,y)$ corresponds to a pair of sequences corresponding to the
	past and future, respectively. The symbol on the $x$ ($y$) axis indicates
	that all points above (to the right of) that symbol have a past (future)
	whose most recent observation is that symbol. Though not marked, further
	proportional subdivisions of each segment of the axes indicate the value of
	the second, third, \emph{etc.} symbols. For instance, one can read from the
	$\mathtt{x+f(x)}$ fractal that any past ending in $\mathtt{f}$ must be
	paired with a future beginning in $\mathtt{(f}$ or $\mathtt{(x}$.
	}
\label{fig:cantor}
\end{figure*}

@@

The intuitive distance metric between probability measures for capturing this
underlying geometry is the Wasserstein metric \cite{Pane19a}. Given two measures
$\mu$ and $\nu$ defined on a metric space $\mathcal{M}$ with metric $d$, the
Wasserstein distance between $\mu$ and $\nu$ is given by:
\begin{align*}
  W(\mu,\nu) = \min_{\pi\in \Gamma(\mu,\nu)}
   \int_{\mathcal{M}\times\mathcal{M}} 
   d(x,y) d\pi(x,y)
   ~,
\end{align*}
where $\Gamma(\mu,\nu)$ is the set of all measures on $\mathcal{M} \times
\mathcal{M}$ whose left and right marginals are $\mu$ and $\nu$, respectively.
It is the minimal cost to ``shift'' the probability mass from one distribution
to match shape of the other.

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=0.45\linewidth]{../plots/even_HCLUST.png}
\includegraphics[width=0.45\linewidth]{../plots/anbn_HCLUST.png}
\includegraphics[width=0.45\linewidth]{../plots/anbncn_HCLUST.png}
\includegraphics[width=0.45\linewidth]{../plots/x+f(x)_HCLUST.png}
\caption{(Upper left to lower right) Clustered Cantor diagrams of the even,
	$\mathtt{a}^n\mathtt{b}^n$, $\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and
	$\mathtt{x+f(x)}$ processes. Zoom for detail. For each, the vertical axis
	shows all pasts of a given length $k$ along with their hierarchically
	clustered dendrogram. $k=8$ for the even, $\mathtt{a}^n\mathtt{b}^n$, and
	$\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$ processes and $k=4$ for the
	$\mathtt{x+f(x)}$ process. For present purposes, the coloring threshold
	was chosen to aid visual interpretation. The lines in each row show the
	empirical distribution of Cantor-embedded futures observed following each
	past. As such, the horizontal axis corresponds exactly to the vertical axis
	of Fig. \ref{fig:cantor}.
	}
\label{fig:hclust}
\end{figure*}

@@

The resulting clustered Cantor plots offer a highly interpretable visualization
of the relationship between pasts and futures, and of the predictive states'
geometry. Each plot, in a certain sense, sorts the columns in the Cantor
fractals of Fig. \ref{fig:cantor} with the white space between columns removed.
For instance, the even process's clustered Cantor plot clearly contains the two
major states, with a third ``transient'' state visible. (The latter corresponds
to the increasingly unlikely event of never seeing a $0$ in a block of length
$n$.) This third state was previously hidden mostly out of view on the
far-right side of the $2$-dimensional Cantor plot of the even process in Fig.
\ref{fig:cantor}.

@@

Similar discernment is seen for the $\mathtt{a}^n \mathtt{b}^n \mathtt{c}^n$
and $\mathtt{x+f(x)}$ processes as well. We draw attention to the manner in
which the presence of a semicolon in pasts from $\mathtt{x+f(x)}$ affects the
comparison of predictions.

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=0.45\linewidth]{../plots/even_MDS.pdf}
\includegraphics[width=0.45\linewidth]{../plots/anbn_MDS.pdf}
\caption{Scatterplots of the first two MDS coordinates of the reconstructed
	predictive states: (Left) Even process. (Right) $\mathtt{a}^n\mathtt{b}^n$
	process. Clusters colored according to the scheme determined by the
	dendrogram in Fig. \ref{fig:hclust} and the label on each cluster describes
	the pattern that uniquely characterizes the pasts in that cluster.
	}
\label{fig:mds}
\end{figure*}

@@

The even process, as in all other cases seen thus far, has two dominant
prediction clusters. These correspond to the predictive states that result
from seeing an even-sized block of $1$s (or, equivalently, no $1$s), and that
result from seeing an odd-sized block of $1$s.

@@

The geometric approach is particularly insightful when computing the
Wasserstein matrix between predictions estimated from Morse-Thue process data.
Recall that the Morse-Thue process is a coarse-graining of the iterated
logistic map $y_{t+1} = ry_t(1-y_t)$ at the critical chaos parameter $r_c
\approx 3.56995$. The resulting stream of $0$s and $1$s is a well-known
instance of high complexity at the ``order-disorder border''. Specifically,
setting parameter $r$ on either side of $r_c$ results in sequences that
can be generated by finite hidden Markov chains. However, at $r_c$ itself the
resulting Morse-Thue process is context-sensitive and therefore requires
infinite predictive states. That is, when it comes to capturing its behavior,
the process is several orders higher in model complexity. It is further up the
Chomsky language hierarchy.

>>>>

\begin{figure*}[ht]
\centering
\includegraphics[width=0.46\linewidth]{../plots/even_CANTOR.png}
\includegraphics[width=0.45\linewidth]{../plots/anbn_CANTOR.png}
\includegraphics[width=0.45\linewidth]{../plots/anbncn_CANTOR.png}
\includegraphics[width=0.45\linewidth]{../plots/x+f(x)_CANTOR.png}
\caption{Cantor plots for the even, $\mathtt{a}^n\mathtt{b}^n$,
	$\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and $\mathtt{x+f(x)}$ processes.
	Each point $(x,y)$ corresponds to a pair of sequences corresponding to the
	past $x$ and future $y$, respectively. The symbol on the $x$ ($y$) axis
	indicates that all points above (to the right of) that symbol have a past
	(future) whose most recent observation is that symbol. Though not marked,
	further proportional subdivisions of each segment of the axes indicate the
	value of the second, third, and so on symbols. For instance, one can read
	from the $\mathtt{x+f(x)}$ fractal that any past ending in $\mathtt{f}$
	must be paired with a future beginning in $\mathtt{(f}$ or $\mathtt{(x}$.
	}
\label{fig:cantor}
\end{figure*}

@@

The intuitive distance metric between probability measures for capturing this
underlying geometry is the Wasserstein metric \cite{Pane19a}. Given two measures
$\mu$ and $\nu$ defined on a metric space $\mathcal{M}$ with metric $d$, the
Wasserstein distance between $\mu$ and $\nu$ is given by:
\begin{align*}
  W(\mu,\nu) = \min_{\pi\in \Gamma(\mu,\nu)}
   \int_{\mathcal{M}\times\mathcal{M}} 
   d(x,y) d\pi(x,y)
   ~,
\end{align*}
where $\Gamma(\mu,\nu)$ is the set of all measures on $\mathcal{M} \times
\mathcal{M}$ whose left and right marginals are $\mu$ and $\nu$, respectively.
It is the minimal cost to ``shift'' the probability mass from one distribution
to match the other's shape.

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=0.49\linewidth]{../plots/even_HCLUST.png}
\includegraphics[width=0.49\linewidth]{../plots/anbn_HCLUST.png}
\includegraphics[width=0.49\linewidth]{../plots/anbncn_HCLUST.png}
\includegraphics[width=0.49\linewidth]{../plots/x+f(x)_HCLUST.png}
\caption{(Upper left to lower right) Clustered Cantor diagrams of the even,
	$\mathtt{a}^n\mathtt{b}^n$, $\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and
	$\mathtt{x+f(x)}$ processes. Zoom for detail. For each, the vertical axis
	shows all pasts of a given length $k$ along with their hierarchically
	clustered dendrogram. $k=8$ for the even, $\mathtt{a}^n\mathtt{b}^n$, and
	$\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$ processes and $k=4$ for the
	$\mathtt{x+f(x)}$ process. For present purposes, the coloring threshold
	was chosen to aid visual interpretation. The lines in each row show the
	empirical distribution of Cantor-embedded futures observed following each
	past. As such, the horizontal axis corresponds exactly to the vertical axis
	of Fig. \ref{fig:cantor}.
	}
\label{fig:hclust}
\end{figure*}

@@

The resulting clustered Cantor plots offer a highly interpretable visualization
of the relationship between pasts and futures and of the predictive states'
geometry. Each plot, in a certain sense, sorts the columns in the Cantor
fractals of Fig. \ref{fig:cantor} with the white space between columns removed.
For instance, the even process's clustered Cantor plot clearly contains the two
major states, with a third ``transient'' state visible. (The latter corresponds
to the increasingly unlikely event of never seeing a $0$ in a block of length
$n$.) This third state was previously hidden mostly out of view on the
far-right side of the $2$-dimensional Cantor plot of the even process in Fig.
\ref{fig:cantor}.

@@

Similar discernment is seen for the $\mathtt{a}^n \mathtt{b}^n \mathtt{c}^n$
and $\mathtt{x+f(x)}$ processes as well. We draw attention to the manner in
which the presence of a semicolon in pasts from the $\mathtt{x+f(x)}$ process
affects the comparison of predictions.

@@

\begin{figure*}[ht]
\centering
\includegraphics[width=0.49\linewidth]{../plots/even_MDS.pdf}
\includegraphics[width=0.48\linewidth]{../plots/anbn_MDS.pdf}
\caption{Scatterplots of the first two MDS coordinates of the reconstructed
	predictive states: (Left) Even process. (Right) $\mathtt{a}^n\mathtt{b}^n$
	process. Clusters colored according to the scheme determined by the
	dendrogram in Fig. \ref{fig:hclust} and the label on each cluster describes
	the pattern that uniquely characterizes the pasts in that cluster.
	}
\label{fig:mds}
\end{figure*}

@@

The even process, as in all other cases seen thus far, has two dominant
prediction clusters. These correspond to the predictive states that result
from seeing an even-sized block of $1$s (or, equivalently, no $1$s) and that
result from seeing an odd-sized block of $1$s. The lone cluster in the middle
corrresponds to a transient state induced by seeing all $1$s and then a $0$.
The latter then synchronizes to right cluster.

@@

The geometric approach is particularly insightful when computing the
Wasserstein matrix between predictions estimated from Morse-Thue process data.
Recall that the Morse-Thue process is a coarse-graining of the iterated
logistic map $y_{t+1} = ry_t(1-y_t), ~y_0 = 0,$ at the critical chaos parameter $r_c
\approx 3.56995$. The resulting stream of $0$s and $1$s is a well-known
instance of high complexity at the ``order-disorder border''. Specifically,
setting parameter $r$ on either side of $r_c$ results in sequences that
can be generated by finite hidden Markov chains. However, at $r_c$ itself the
resulting Morse-Thue process is context-sensitive and therefore requires
infinite predictive states. That is, when it comes to capturing its behavior,
the process is several orders higher in model complexity. It is further up the
Chomsky language hierarchy.
